\section{Introduction}

	This document contains the assignments for the Stratego
	Tutorial.  The purpose of the assignments is to define several
	transformations on Pico programs. Pico is a small imperative
	language with nested function declarations.

\section{Syntax}

 \subsection{Examples}

   \subsubsection{Fib}

	\inputverbatim{../xmpl/fib.pico}

 \subsection{Syntax Definition}

	\inputverbatim{../syn/pico.def}

\section{Abstract Syntax}

	\inputverbatim{../spec/pico.r}

\section{Pretty-Printer}

	\inputverbatim{../spec/pico-pp.r}

\section{Lab Environment}

	\begin{itemize}

	\item There is a guest account for each of you

	\item Add to your path: /projects/stratego/xt/bin

	\item To use the emacs mode put the following in
	\verb|~/.emacs|:
        \verb|(load "/projects/stratego/xt/share/stratego/stratego.el")|

	\item Set up your work directory in
	/projects/stratego/tutorial/firstname (you don't have enough
	quota to work comfortably in your home directory)

	\end{itemize}

\section{Assignments}

 \subsection{Algebraic Simplification (1)}

	Use the well-known algebraic laws about arithmetic to simplify
	arithmetic expressions.

 \subsection{Three Address Code}

	Normalize all expressions to the form \verb|x := y + z|, i.e.,
	no complex expressions are used.

 \subsection{Side-Effect Free}

	Remove all side-effects from expressions, i.e., 
\verb|x + (x := y - z)| 
	should become 
\verb|x + (x := y - z; x)|,
	which should transform to
\verb|(var a; a := x; x := y - z; a + x)|

 \subsection{Desugaring}

	Introduce a for loop and express it in terms of a while

 \subsection{Algebraic Simplification (2)}

	Simplify logical expressions using Boolean algebra.

 \subsection{Mutification}

	Replace move statements (:=) by operator-move statement (+=)
	whenever possible. For example, replace \verb|x := x + e| with
	\verb|x += e|.

 \subsection{Inlining}

	Replace a function call by its body.

\subsection{Evaluation}
	
	Write an Interpreter for Pico.
