
% Copyright (C) 1998, 1999 Eelco Visser <visser@acm.org>
% 
% This program is free software; you can redistribute it and/or modify
% it under the terms of the GNU General Public License as published by
% the Free Software Foundation; either version 2, or (at your option)
% any later version.
% 
% This program is distributed in the hope that it will be useful,
% but WITHOUT ANY WARRANTY; without even the implied warranty of
% MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
% GNU General Public License for more details.
% 
% You should have received a copy of the GNU General Public License
% along with this program; if not, write to the Free Software
% Foundation, Inc., 59 Temple Place - Suite 330, Boston, MA
% 02111-1307, USA.

\chapter{Further Reading}

\subsection{Extending Your Understanding of Stratego}

	\begin{abstract}

	This tutorial only scratches the surface of Stratego
	specification. To extend your understanding of Stratego refer
	to the other existing documentation and bug me with requests
	to extend this tutorial.
	
	\end{abstract}

	\paragraph{Other Documentation}

	Refer to \emph{The Stratego Reference Manual} for an overview
	of all the language constructs.

	Refer to \emph{The Stratego Library} for a large number of
	standard strategies. The modules in the library are self
	documenting and often contain examples of the application of
	the strategies.

	Refer to \emph{The Stratego Compiler} for an overview of the
	implementation of Stratego and all the details of the
	compilation process.

	\paragraph{Publications}

	There are a number of publications on the design,
	implementation and use of Stratego. See the bibliography for
	references. While these publications do not always reflect the
	current state of the language, they do reflect its spirit.

	\paragraph{Topics for Future Inclusion in this Tutorial}

	Here are some topics that I consider for inclusion in this
	tutorial. Which ones I tackle first will depend on popular
	demand and my mood.
	
	\begin{itemize}

	\item Discussion of (advanced) language constructs

	\item Primitives

	\item Architectures for program transformation

	\item Connecting Stratego components to other systems

	\end{itemize}


\clearpage